Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 paper.tex
summary done, please check.  added middle initial

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Vertical shear instability}
\shortauthors{M-K. Lin, A. Youdin}

@@

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI.
   We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial  
  powerlaw temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to PPD models 
  where $t_c$ is determined by the opacity of dust grains.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

\section{Introduction}\label{intro}
Understanding how disks transport mass and angular momentum underlies 
many problems in astrophysics, including star and planet formation  
\citep[][]{lyndenbell74, armitage10}.  The turbulence associated with many transport mechanisms
is particularly important for dust evolution and planetesimal formation \citep{yl07, chiang10}.

@@

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars (\citealp{goldreich67}, hereafter \citetalias{goldreich67}, \citealp{fricke68}).   
The VSI arises when vertical shear, i.e. \ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Vertical shear instability}
\shortauthors{M-K.\ Lin, A.\ N.\ Youdin}

@@

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically
%global linear calculations with thermal relaxation and applications
%to the Solar Nebula}  
\author{Min-Kai Lin \& Andrew N Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, magnetic transport mechanisms that
  operate in warmer disks are less effective.  
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks (PPDs).  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong
  stabilizing influence in cold disks, where  
  heating is dominated by external irradiation.  Rapid cooling, via
  radiative losses, reduces the effective buoyancy 
  and allows the VSI to operate.  In this paper, we quantify the
  cooling timescale, $t_c$, needed for growth of the VSI.
   We perform a linear analysis of the VSI with cooling in
  vertically global and radially local disk models. 
  For irradiated disks, we find that the VSI is most vigorous for
  rapid cooling with $t_c < \OmK^{-1}  h |q| / (\gamma -1)$ 
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's
  aspect ratio, $ h \ll 1$; the radial  
  powerlaw temperature gradient, $q$; and the adiabatic index,
  $\gamma$.  For longer cooling times, the VSI  
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to PPD models 
  where $t_c$ is determined by the opacity of dust grains.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.
  Growth is suppressed by long cooling times both in the opaque  
  inner disk and the optically thin outer disk.  A reduction in the
  dust opacity by a factor of 10 increases cooling times  
  enough to quench the VSI at all disk radii.  Thus the formation of
  solid protoplanets, a sink for dust grains, can impede the VSI. 
\end{abstract}

\section{Introduction}\label{intro}
Understanding how disks transport mass and angular momentum underlies 
many problems in astrophysics, including star and planet formation  
\citep[][]{armitage10}.  The turbulence associated with many transport mechanisms
is particularly important for dust evolution and planetesimal formation \citep{yl07, chiang10}.

@@

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed by the positive radial specific angular
momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, urpin03, nelson13}, hereafter \citetalias{nelson13}).  The
basic mechanism of the VSI  in disks was first 
identified in the context of differentially rotating stars (\citealp{goldreich67}, hereafter \citetalias{goldreich67}, \citealp{fricke68}).   
The VSI arises when vertical shear, i.e.\ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces. Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in astrophysical disks, except at special locations like the
midplane.

### 2015-05-08 summary.tex
re-shuffled 6.4, updated fig 2, poster files update
summary done, please check.  added middle initial

\subsection{Convective overstability}
The VSI bears some similarity to the `convective overstability'
 (CONO, \citealp{klahr14, lyra14}).   Both instabilities rely on thermal relaxation to 
overcome the stability of adiabatic perturbations, i.e.\ the Solberg-Hoiland
criteria (see \S\ref{solberg}).

@@

The sign of $N_r^2$ depends on disk parameters.  To demonstrate that the required parameters 
are somewhat extreme, we consider the midplane of vertically isothermal disks with 
$\Sigma \propto r^n$ so that $n = p+q/2 + 3/2$. Using Eqs. \ref{dSdr} and \ref{Nr} with $\gamma = 1.4$,
$N_r^2 < 0$ requires $n> 3(1/2 + q)$.  For standard disk temperature laws this requirement implies
a flat or rising surface density profile, e.g.\ $n = 3/14$ for $q = -3/7$ or $n = 0$ for $q = -1/2$.
Thus CONO is most like to operate at special locations, like the outside edges of 
disk gaps or holes and/or shadowed regions where $q < -1/2$.

Perhaps due to these physical differences, CONO operates in different regimes of parameter
space than the VSI.  CONO grows best for longer wavelengths $\khat \lesssim 1$ and longer 
cooling times $\beta \sim 1$.

@@

\subsection{Radiative transfer} 
Our relatively simple treatment of cooling with an idealized dust opacity
could certainly be generalized in future works.  In hotter disks, gas phase opacities
must be considered.   In cold disks there are many choices for a dust opacity which 
varies with grain sizes and compositions. Dust properties could change the viability 
of the VSI for better or worse.  The radiative transfer itself
could be calculated with higher levels of sophistication, as is already being done in
numerical simulations \citep{stoll14}.

\section{Summary and discussion}\label{summary}
In this paper we study the vertical shear instability (VSI) with a focus on 
the radiative cooling requirements.  In turn we address the viability of the VSI 
as an angular momentum transport mechanism in protoplanetary disks (PPDs).

Our linear axisymmetric analysis of the VSI 
considers (uniquely to our knowledge) finite cooling times in a vertically global model.  
Our main analytical finding, which we confirm numerically, is the critical cooling 
timescale above which VSI growth is suppressed, Eqs.\ \ref{prelim_bcrit} and \ref{iso_vsi_cond}.
Short cooling times are needed to weaken the stabilizing effects of vertical buoyancy, thereby allowing 
the free energy in vertical shear to drive instability.

Our main focus is irradiated, vertically isothermal disks which have strong vertical buoyancy.
The critical cooling time is thus short, shorter than the orbital time by a factor of the disk
aspect ratio.  This finding is consistent with, and helps explain, the results of 3D numerical simulations \citepalias{N13}.

In applying our results to PPDs, we pay particular attention to the transition from 
optically thick radiative diffusion to optically thin Newtonian cooling.  The largest obstacle to VSI occurs in 
high density inner disk, $\lesssim 1$ -- 5 AU, where radiative diffusion times are slow.  Shorter wavelength disturbances
 cool faster, though only if they remain optically thick.  In the inner disk cooling can only be fast enough
 if wavelengths are too short to drive significant transport.  The best hope for the VSI in inner disk is a low surface density.

...

In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile,  
$\p_z\Omega\neq 0$. Vertical shear is generally
expected in astrophysical disks due to baroclinicity ($\nabla
P\times\nabla \rho \neq 0$). This may be due
to a radial variation in the disk temperature, or more generally in
the disk entropy. However, astrophysical disks are also typically
sub-adiabatically stratified for stability against vertical
convection ($N_z^2\geq 0$). In this case, the VSI requires a short
thermal or cooling timescale $t_c$ in order to operate.

Physically, a short thermal timescale is needed to reduce the  
stabilizing effect of buoyancy forces. Because vertical shear is weak
in thin disks, very rapid cooling, corresponding to $t_c\Omega\ll 1$, 
is generally needed for the VSI, otherwise buoyancy is strongly
stabilizing. Our primary goal was to quantify this requirement,
thereby provide a simple way to assess whether or not the VSI is
relevant in realistic disks.

To do so, we generalized previous vertically global, radially local linear 
calculations of the VSI \citepalias{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk 
temperature to its initial value on a timescale
$t_c$. 
% This `thermal relaxation' model allows
% the thermodynamic response of the disk to be controlled by varying
% $t_c$, and has been employed in previous numerical simulations of the
% VSI \citepalias{nelson13}.  
%In accordance with the above studies, which show that the VSI occurs for 
%perturbations with small radial lengthscales, we applied the 
%radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as this is the
relevant case for protoplanetary disks (PPDs) passively 
irradiated by its central star \citep{chiang97}. It also permits a
simplified analytical discussion by extending the 
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) cooling. We obtained explicit solutions 
describing the VSI with thermal relaxation in the thin disk limit.

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\OmK^{-1}, 
\end{align*}
below which the VSI operates effectively. 
%While $t_\mathrm{crit}$ was
%derived for the lowest order VSI mode --- the `fundamental  
%corrugation mode' --- we find from numerical calculations that
%$t_\mathrm{crit}$ can be used to characterize the general
%stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have thus provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is 
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

We applied our linear models to assess the applicability of the VSI in
PPDs. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on radial lengthscales signficantly shorter
than the disk scaleheight. %However, such small-scale perturbations may 
%be subject to viscous decay in a real disk.

Although $t_c$ is not vertically constant in the 
MMSN as assumed in our analysis, we find that evaluating the condition
$t_c< t_\mathrm{crit}$ at the disk midplane gives an indication of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a practical first step in estimating the relevance of the VSI in
realistic astrophysical disks without resorting to linear
calculations or non-linear numerical simulations.

>>>>

\subsection{Convective overstability}
The VSI bears some similarity to the `convective overstability'
 (CONO, \citealp{klahr14, lyra14}).   Both instabilities rely on thermal relaxation to 
overcome the stability of disks to adiabatic perturbations, i.e.\ the Solberg-Hoiland
criteria (see \S\ref{solberg}).

@@

The sign of $N_r^2$ depends on disk parameters.  To demonstrate that the parameters needed for $N_r^2 <0$
are somewhat extreme, or at least non-standard, we consider the midplane of vertically isothermal disks with 
$\Sigma \propto r^n$ so that $n = p+q/2 + 3/2$. Using Eqs. \ref{dSdr} and \ref{Nr} with $\gamma = 1.4$,
$N_r^2 < 0$ requires $n> 3(1/2 + q)$.  For standard disk temperature laws, this requirement implies
a flat or rising surface density profile, e.g.\ $n = 3/14$ for $q = -3/7$ or $n = 0$ for $q = -1/2$.
Since standard $\Sigma$ profiles decline with radius, CONO is most like to operate at 
special locations, like the outside edges of disk gaps or holes and/or shadowed regions where $q < -1/2$.

Perhaps due to these physical differences, CONO operates in different regimes of parameter
space than the VSI.  The CONO grows best for longer wavelengths $\khat \lesssim 1$ and longer 
cooling times $\beta \sim 1$.

@@

\subsection{Radiative transfer} 
Our relatively simple treatment of cooling with an idealized dust opacity
could certainly be generalized in future works.  In hotter disks, gas phase opacities
must be considered.   In cold disks, there are many choices for the dust opacity, which 
varies with grain sizes and compositions. Changing dust properties would alter 
the viability of the VSI for better or worse.  The radiative transfer itself
could be calculated with higher levels of sophistication, as is already being done in
numerical simulations of the VSI \citep{stoll14}.

\section{Summary and discussion}\label{summary}
In this paper we study the vertical shear instability (VSI) with a focus the 
role of radiative cooling.  In turn we assess the viability of the VSI 
as an angular momentum transport mechanism in protoplanetary disks (PPDs).

Our linear, axisymmetric analysis of the VSI 
considers (uniquely to our knowledge) finite cooling times in a vertically global model.  
In order for vertical shear to drive the VSI, short cooling times are needed to weaken 
the stabilizing effects of vertical buoyancy. Our main analytical finding, which we confirm numerically, 
is the critical cooling  timescale above which VSI growth is suppressed, Eqs.\ \ref{prelim_bcrit} and \ref{iso_vsi_cond}.

Our main focus is irradiated, vertically isothermal disks which have strong vertical buoyancy.
The critical cooling time is thus short, shorter than the orbital time by a factor of the disk
aspect ratio.  This finding is consistent with, and helps explain, the results of 3D numerical simulations \citepalias{nelson13}.

In applying our results to PPDs, we pay particular attention to the transition from 
optically thick radiative diffusion to optically thin Newtonian cooling.  The largest obstacle to VSI occurs in 
high density inner disk, $\lesssim 1$ -- 5 AU, where radiative diffusion times are slow.  Shorter wavelength disturbances
 cool faster, so long as they remain optically thick.  In the inner disk, however, our VSI cooling criterion requires 
 wavelengths that are too short to drive significant transport.  The best hope for the VSI in inner disk appears to 
 be a low surface density, which speeds radiative diffusion by lengthening the photon mean free path.  This option naturally begs 
 the question of what accretion mechanism lowered the surface density in the first place.

 In the outer disk, the VSI tends to cool in the optically thin limit.  The main issue is whether the opacity 
 is high enough for optically thin cooling to be sufficiently fast.  Our standard opacity assumes a Solar abundances of small dust.
 In this case, the VSI can operate from $\sim 5$ -- 100 AU.  A factor of 10 reduction in the opacity, for instance by locking
 small dust into planetesimals and planetary cores \citep{youdin13}, makes cooling times too slow for VSI growth.  
 In this case cooling is too slow not just in the outer disk, but into  $< 1$ AU.

 An enhanced opacity makes optically thin cooling faster and radiative diffusion slower.  This shift favors VSI growth 
 in the outer disk at the expense of the inner disk, pushing the inner limit of VSI growth further out.
 The standard choice -- corresponding to Solar abundances in a standard MMSN disk -- allows the VSI 
 to grow over the widest range of relevant disk radii.

 Our detailed study of the spectrum of VSI modes confirms that some artificial `surface modes' are 
 triggered by imposed vertical boundaries \citepalias{nelson13, barker15}.  Domain size
 convergence studies are thus essential.  Fortunately, our results show that longer cooling times 
  stifle the growth of surface modes.  Thus, at least in some cases, more realistic radiative transfer also produces more reliable dynamics.

 The VSI deserves further study as a viable mechanism to drive at least low levels of accretion in cold disks.
